Sparse Function-space Representation \\ of Neural Networks

### 2023-05-15 paper/main.tex
updated rl section

% by linearising a neural network around the Maximum a Posteriori (MAP) weights.
% a GP posterior -- can be obtained
%
In supervised learning, given a data set $\dataset = \{(\vx_{i} , \vy_{i})\}_{i=1}^{n}$, with input $\vx_i \in \inputDomain$ and output $\vy_i \in \outputDomain$ pairs, the weights $\weights \in \R^{P}$ of a neural network, $f_\mathbf{w} : \inputDomain \to \outputDomain$, are usually trained to minimize the (regularized) empirical risk,
%
\begin{equation} \label{eq-empirical-risk}
  \weights^{*} = 
  \arg \min_{\weights} \mathcal{L}(\dataset,\weights) =
  \arg \min_{\weights} \textstyle\sum_{i=1}^{n} \ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i) + \delta \mathcal{R}(\weights).
\end{equation}
%
If $\ell(f_\weights(\vx_{i}), \vy_i) = -\log(p(\vy \mid f_\weights(\vx_{i}))$ and the regularizer $\mathcal{R}(\weights) = \frac{1}{2}\|\weights\|^{2}_2$, then we can view \cref{eq-empirical-risk} as the maximum {\it a~posteriori} (MAP) solution to a Bayesian objective, where the regularization weight takes the role of a prior precision parameter, \ie, $p(\vw) = \Norm(\vzeros, \delta^{-1} \MI)$.
%Bayesian inference offers a principled approach to quantifying uncertainty in neural networks.
%The goal is to find the posterior over the weights ${p(\vw \mid \vy) \propto p(\vy \mid f_{\weights}(\vx)) \, p(\weights)}$ as it represents our belief in the parameters after combining data $\dataset$ with our prior $p(\vw)$.
%Although the true posterior $p(\vw \mid \dataset)$ is intractable given the non-linearities of the neural network, one can resort to sampling techniques such as Hamiltonian Monte-Carlo (HMC).
%However, for most applications their high computational costs make them impractical so we need approximate inference techniques.
%In \cref{fig:banana}, we show a qualitative example of the induced function-space posterior obtained with our method compared to the posterior obtained through HMC sampling on the Banana toy dataset. 
The posterior over the weights ${p(\vw \mid \dataset) \propto p(\vy \mid f_{\weights}(\vx)) \, p(\weights)}$ is generally not available in closed form. Sampling methods that characterize the posterior with a finite set of samples in weight-space---such as the Hamiltonian Monte Carlo (HMC) baseline used in \cref{fig:banana} (right)---are general-purpose, but comuptationally heavy and impractical for downstream applications.

@@

\subsection{\our for Continual Learning}
% Definition of CL -> a sequence of tasks 
In the Continual learning setting, the training is divided into $T$ tasks, each with its training data set $\dataset_t = \{(\mathbf{x}_{i}, \mathbf{y}_{i})\}_{i=1}^{n_t}$, which after the task is discarded and cannot be accessed in the following stages of the training. Rehearsal and function regularization-based models keep a subset of the training data for each task to help the model alleviate forgetting. In the same way, after each task, we can use our \our to build a compact representation of the neural network and then exploit this representation to build a functional regularizer to add to the training loss for the subsequent tasks. Recently proposed GP-based regularizers \cite{ pan2020continual, rudner2022continual} have shown better performances than weight space equivalents getting state-of-the-art on CL benchmarks. However, they are limited to use subset approximations of the posterior.
%, which in practice means randomly selecting some data computing its posterior at only those points and then using that as the regularizer. 
A more principled approach would be to sparsify the resultant GP into a low-rank approximation to ensure information coming from the entire task dataset is not lost.%, essentially what the sparse dual variables do.   \todo{this term is replacing the missing data}
% Where the regularizer comes from
At the end of the training for each task $t$, we compute the Laplace-GGN approximated kernel function for the current model and randomly select $m$ inducing points. Given the issues described in the previous sections, related to computing the full covariance posterior $\MKxx$,\todo{this should be $\kappa(\MX_t, \MX_t)$}, we can exploit the dual parameterization to efficiently encode the information from the $\dataset_t$ using the dual parameterization, as follows, 
\begin{equation}
 	\bar{\MB}^{-1}_t = \MKzz^{-1} \vbeta_\vu \MKzz^{-1} \in \R^{m \times m}, 
 	\quad 
 	\vbeta_{\vu} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
 	% \quad \forall t \in [1, T]
 \end{equation}
 where $\MKzz$ and $\vkzi$ are the kernel function computed on the inducing points $\MZ_t \in \R^{m \times D}$ for task $t$.

@@

After each task, we update $\mathcal{M}$, and we can construct a regularizer term, which aims to reduce the drift of the current model from diverging on the inducing points $\MZ_s$ for all previous tasks $ s \in \left[ 1, t-1 \right]$. The resulting regularizer can be computed as follows
%\begin{equation}
%	\mathcal{R}_\textit{SFR}(\weights, \mathcal{M}) = \sum_{s=1}^{t-1} \frac{1}{m} \left[\left(f_{\weights}(\MZ_{s}) - f_{\weights_{s}}(\MZ_s) \right)^\T \bar{\MB}^{-1}_{s} \left(f_{\weights}(\MZ_{s}) - f_{\weights_{s}}(\MZ_s) \right) \right].
%\end{equation}

\begin{equation}
	\mathcal{R}_\textit{SFR}(\weights, \mathcal{M}) = \sum_{s=1}^{t-1} \frac{1}{m} 
	\left\lVert 
	f_{\weights}(\MZ_{s}) - \vu_t % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}}.
\end{equation}
%\todo{choose one option}
% where $f_\vw(\MZ_{s}) \in \R^m$ are the outputs of neural networks with the current weights $\vw$ and stored function outputs of the neural network at the MAP optimum for task $s$, i.e., each $i$th entry is $f_\cdot(\vz_i) \in \R$.
%\begin{equation}
%	\argmin_{\weights} \sum_{i=1}^{n} \ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i) + \delta \mathcal{R}(\weights) + \tau \mathcal{R}_\textit{SFR}(\weights)
%\end{equation}
Then, when the training on the successive task, the new objective to minimize takes the following form
\begin{equation}
	\weights_t^* = \argmin_{\weights} \mathcal{L}(\mathcal{D}_t, \weights) + \tau \mathcal{R}_\textit{SFR}(\weights, \mathcal{M})
\end{equation}

@@

\subsection{Reinforcement learning}
A key challenge in RL is balancing the trade-off between exploration and exploitation.
That is, should an agent select actions that it knows will lead to a high reward (exploitation), or should it
hope to discover new actions which lead to high reward (exploration)?
In this section, we show that \our's uncertainty estimates can be used to balance this trade-off in a model-based RL algorithim,
demonstrating the quality of our uncertainty estimates.
% In this section, we demonstrate  that our method's principled uncertainty estimates can be used to balance this trade-off in a model-based RL algorithim.
We refer the reader to \cref{app:rl} for an overview of the problem and details of our experiments.

Cart pole swing up task in MuJoCo \cite{todorov2012mujoco}
We compare our method with deep deterministic policy gradient (DDPG) \cite{lillicrapContinuousControlDeep2016} -- a model-free RL baseline --
which is sample inefficient.
We test our model's uncertainty estimates by comparing it to using Laplace-GGN with GLM predictions, using an ensemble of neural networks
(with different initialisations),
To ensure a fair comparison, we use the same MLP architecture/training scheme and we keep the same model-based RL algorithm.
See \cref{app:rl} for more details of our experiments.

We also compare to deep deterministic policy gradient (DDPG) \cite{lillicrapContinuousControlDeep2016} -- a model-free RL baseline --
which we expect to be much more sample inefficient.

\cref{fig:rl} shows training curves

@@

\section{\our for continual learning}
\label{app:cl}
\subsection{Derivation of the regularizer term}
An alternative view of the dual variables is that they parameterize approximate likelihoods, as shown in \citep{adam2021dual, khan2017conjugate}. We can rewrite the GP posterior for $q(\vu)$ in terms of the following approximate normal distributions:
\begin{equation}
 q(\vu) \propto \Norm(0, \MKzz) \prod_{i=1}^n \exp \! \left(-\frac{\tilde{\beta}_i}{2}(\tilde{y}_i - \va_i^\top \vu)^2 \right) = \Norm(0, \MKzz) \, \exp\left((\bar{\vy} - \vu)^\top \bar{\MSigma}^{-1}(\bar{\vy} - \vu)\right)  
\end{equation}
The first normals are the approximate likelihoods where $\va_i^\top = \vkzi^\top \MKzz^{-1}$ and $\tilde{y}_i = \hat{\alpha}_i + f_{\vw^*}(\vx_i)\hat{\beta}$,  replacing $\va_i^\top \vu$ with $\vx_i^{\top}\vw$ and we can see the similarity between the sparse GP and a linear model where the kernel matrix determines the prior. To arrive at the MVN Gaussian, we need to rearrange the quadratic terms to match the prior sufficient statistics we present a derivation in [appendix]. Performing conjugate summation of prior and likelihood, we would arrive at the sparse GP posterior at the inducing points. The correlated covariance structure $\MSigma$ and the corresponding $\tilde{\vy}$ are simply a different form of the sparse dual variables,
\begin{equation}
\quad \bar{\MSigma} =  \MKzz \vbeta_\vu^{-1} \MKzz \quad \bar{\vy} = \MKzz \vbeta_{\vu}\valpha_{\vu}.
\end{equation}
Written as above, we can see that the spare dual variables can be interpreted as a sparse MVN representation. We now show how we use this for a regularizer in continual learning with more details in [appendix].

>>>>

% by linearising a neural network around the Maximum a Posteriori (MAP) weights.
% a GP posterior -- can be obtained

In supervised learning, given a data set $\dataset = \dataset = \{(\mathbf{x}_{i} , \mathbf{y}_{i})\}_{i=1}^{n}$, with input $\mathbf{x}_i \in \inputDomain$ and output $\mathbf{y}_i \in \outputDomain$ pairs, the weights $\weights \in \R^{P}$ of a neural network, $f_\mathbf{w} : \inputDomain \to \outputDomain$, are usually trained to minimize the (regularized) empirical risk,
%
\begin{equation} \label{eq-empirical-risk}
  \weights^{*} = 
  \arg \min_{\weights} \mathcal{L}(\dataset,\weights) =
  \arg \min_{\weights} \sum_{i=1}^{n} \ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i) + \delta \mathcal{R}(\weights).
\end{equation}
%
If $\ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i) = -\log(p(\mathbf{y} \mid f_\weights(\mathbf{x}_{i}))$ and the regularizer $\mathcal{R}(\weights) = \frac{1}{2}\|\weights\|^{2}_2$, then we can view \cref{eq-empirical-risk} as the maximum {\it a~posteriori} (MAP) solution to a Bayesian objective, where the regularization weight takes the role of a prior precision parameter, \ie, $p(\vw) = \Norm(\vzeros, \delta^{-1} \MI)$.

%
% \subsection{Linearization gives rise to a Gaussian process}
% \label{sec:nn2gp}
%

\textbf{Uncertainty quantification} \todo{Move this section to the related work}
Bayesian inference offers a principled approach to quantifying uncertainty in neural networks.
The goal is to find the posterior over the weights ${p(\vw \mid \vy) \propto p(\vy \mid f_{\weights}(\vx)) p(\weights)}$ as
it represents our belief in the parameters after combining data $\dataset$ with our prior $p(\vw)$.
Although the true posterior $p(\vw \mid \dataset)$ is intractable given the non-linearities of the neural network, one can resort to sampling techniques such as Hamiltonian Monte-Carlo (HMC).
However, for most applications their high computational costs make them impractical so we need approximate inference techniques.
In \cref{fig:banana}, we show a qualitative example of the induced function-space posterior obtained with our method compared to the posterior obtained through HMC sampling on the Banana toy dataset.

@@

\subsection{\our for Continual Learning}
% Definition of CL -> a sequence of tasks 
In the Continual learning setting, the training is divided into $T$ tasks, each with its training data set $\dataset_t = \{(\mathbf{x}_{i}, \mathbf{y}_{i})\}_{i=1}^{n_t}$, which after the task is discarded and cannot be accessed in the later stages of the training. However, rehearsal and function regularization-based models keep a subset of the training data for each task to help the model alleviate forgetting. After each task, we can use our \our to build a compact representation of the neural network and then use that as an additional term to the training loss to regularize the model. Recent success has shown GP-based regularizers perform better than weight space equivalents getting state-of-the-art on typical CL benchmarks \cite{ pan2020continual, rudner2022continual}. The limitation of such approaches is that they must use subset approximations of the posterior, which in practice means randomly selecting some data computing its posterior at only those points and then using that as the regularizer. A more principled approach would be to sparsify the data into a low-rank approximation to ensure information wasn't lost, essentially what the sparse dual variables do.   \todo{this term is replacing the missing data}

An alternative view of the dual variables is that they parameterize approximate likelihoods, as shown in \citep{adam2021dual, khan2017conjugate}. We can rewrite the GP posterior for $q(\vu)$ in terms of the following approximate normal distributions:
\begin{equation}
 q(\vu) \propto \N(0, \MKzz) \prod_{i=1}^n e^{-\frac{\tilde{\beta}_i}{2}(\tilde{y}_i - \va_i^\top \vu)^2 } = \N(0, \MKzz) \, e^{(\bar{\vy} - \vu)^\top \bar{\MSigma}^{-1}(\bar{\vy} - \vu)}  
\end{equation}
The first normals are the approximate likelihoods where $\va_i^\top = \vkzi^\top \MKzz^{-1}$ and $\tilde{y}_i = \hat{\alpha}_i + f_{\vw^*}(\vx_i)\hat{\beta}$,  replacing $\va_i^\top \vu$ with $\vx_i^{\top}\vw$ and we can see the similarity between the sparse GP and a linear model where the kernel matrix determines the prior. To arrive at the MVN Gaussian, we need to rearrange the quadratic terms to match the prior sufficient statistics we present a derivation in [appendix]. Performing conjugate summation of prior and likelihood, we would arrive at the sparse GP posterior at the inducing points. The correlated covariance structure $\MSigma$ and the corresponding $\tilde{\vy}$ are simply a different form of the sparse dual variables,
\begin{equation}
\quad \bar{\MSigma} =  \MKzz \vbeta_\vu^{-1} \MKzz \quad \bar{\vy} = \MKzz \vbeta_{\vu}\valpha_{\vu}.
\end{equation}
Written as above, we can see that the spare dual variables can be interpreted as a sparse MVN representation. We now show how we use this for a regularizer in continual learning with more details in [appendix].

% Where the regularizer comes from
At the end of the training for each task $t$, we compute the Laplace-GGN approximated kernel function for the current model and randomly select $m$ inducing points. Given the issues described in the previous sections, related to computing the full covariance posterior $\MKxx$,\todo{this should be $\kappa(\MX_t, \MX_t)$}, we can exploit the dual parameterization to efficiently encode the information from the $\dataset_t$ using the dual parameterization, as follows, 
\begin{equation}
 	\bar{\MB}^{-1}_t = \MKzz^{-1} \vbeta_\vu \MKzz^{-1} \in \R^{m \times m}, 
 	\quad 
 	\vbeta_{\vu} =  \sum_{i \in \dataset_t} \vkzi \,\hat{\beta}_{i} \, \vkzi^{\T} ,    
 	% \quad \forall t \in [1, T]
 \end{equation}
 where $\MKzz$ and $\vkzi$ are the kernel function computed on the inducing points $\MZ_t \in \R^{m \times D}$ for task $t$.

@@

After each task, we update $\mathcal{M}$, and we can construct a regularizer term, which aims to reduce the drift of the current model from diverging on the inducing points $\MZ_s$ for all previous tasks $ s \in \left[ 1, t-1 \right]$. The resulting regularizer can be computed as follows
\begin{equation}
	\mathcal{R}_\textit{SFR}(\weights, \mathcal{M}) = \sum_{s=1}^{t-1} \frac{1}{m} \left[\left(f_{\weights}(\MZ_{s}) - f_{\weights_{s}}(\MZ_s) \right)^\T \bar{\MB}^{-1}_{s} \left(f_{\weights}(\MZ_{s}) - f_{\weights_{s}}(\MZ_s) \right) \right].
\end{equation}

\begin{equation}
	\mathcal{R}_\textit{SFR}(\weights, \mathcal{M}) = \sum_{s=1}^{t-1} \frac{1}{m} 
	\left\lVert 
	f_{\weights}(\MZ_{s}) - \vu_t % f_{\weights_{s}}(\MZ_s)
	\right\rVert_{\bar{\MB}^{-1}_{s}}.
\end{equation}
\todo{choose one option}
% where $f_\vw(\MZ_{s}) \in \R^m$ are the outputs of neural networks with the current weights $\vw$ and stored function outputs of the neural network at the MAP optimum for task $s$, i.e., each $i$th entry is $f_\cdot(\vz_i) \in \R$.

%\begin{equation}
%	\argmin_{\weights} \sum_{i=1}^{n} \ell(f_\weights(\mathbf{x}_{i}), \mathbf{y}_i) + \delta \mathcal{R}(\weights) + \tau \mathcal{R}_\textit{SFR}(\weights)
%\end{equation}
Then, when the training on the successive task, the new objective to minimize takes the following form
\begin{equation}
	\weights_t^* = \argmin_{\weights} \mathcal{L}(\mathcal{D}_t, \weights) + \tau \mathcal{R}_\textit{SFR}(\weights, \mathcal{M})
\end{equation}

@@

\subsection{Reinforcement learning}
We now demonstrate the quaility of \our's uncertainty estimates by showing that it can improve sample efficiency in RL when used in conjuction with an
uncertainty-guided exploration strategy.
Balancing the trade-off between exploration and exploitation is a key challenge in RL \cite{sutton2018reinforcement}.
Should an agent select actions that it knows will lead to high reward (exploitation), or actions which it has not select before in the
hope to discover new high reward actions (exploration)?

Prior work has used GP dynamic models \cite{deisenrothPILCO2011,kamtheDataEfficient2018}, ensembles of neural
networks \cite{curiEfficient2020,chuaDeepReinforcementLearning2018}
variational inference \cite{galImproving2016,houthooftVIME2017}

% In this section, we show that \our's uncertainty estimates can be used to balance this trade-off in a model-based RL algorithim,
% demonstrating the quality of our uncertainty estimates.

% A key challenge in RL is balancing the trade-off between exploration and exploitation.
% That is, should an agent select actions that it knows will lead to a high reward (exploitation), or should it
% hope to discover new actions which lead to high reward (exploration)?
% In this section, we show that \our's uncertainty estimates can be used to balance this trade-off in a model-based RL algorithim,
% demonstrating the quality of our uncertainty estimates.
% In this section, we demonstrate  that our method's principled uncertainty estimates can be used to balance this trade-off in a model-based RL algorithim.
We refer the reader to \cref{app:rl} for an overview of the reinforcement learning problem and more details of our experiments.

% It is an inherently unstable and underactuated mechanical system.

by using it to learn a dynamics model and
evaluate the
We test \our by using it to learn a dynamics model
our method on the cart pole swing up task in MuJoCo \cite{todorov2012mujoco}, a classic benchmark for nonlinear control, see \cref{fig:rl}.
The goal is to swing the pole up and balance it around the upward position.
We increase the  difficulty of exploration in this environment by using a sparse reward function.

We compare our method to a Laplace-GGN with GLM predictions (red), an ensemble of neural networks (green) and a basic MLP wih no uncertainty (magenta).
To ensure a fair comparison, we use the same MLP architecture/training scheme and use them in the same model-based RL algorithm, detailed in \cref{app:rl}.
% See \cref{app:rl} for more details of our experiments.
We also compare to deep deterministic policy gradient (DDPG) \cite{lillicrapContinuousControlDeep2016}, a model-free RL algorithim (yellow).
% As expected DDPG is the least sample inefficient
The training curves in \cref{fig:rl} show that \our's uncertainty estimates are good for exploration as it converges in fewer episodes.

% We compare our method with deep deterministic policy gradient (DDPG) \cite{lillicrapContinuousControlDeep2016} -- a model-free RL baseline --
% which is sample inefficient.

@@

\section{Sparse functional regularisation in continual learning}
\label{app:cl}
